---
title: "Energy-I-Score: First Steps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example_IScore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette provides a basic introduction to using the `miceDRF` package.  
To install the latest development version directly from GitHub, run the following code in your R console:

```{r, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
devtools::install_github("KrystynaGrzesiak/miceDRF")
```

After installation, load the package and set a random seed to ensure reproducibility:

```{r, message=FALSE, warning=FALSE}
library(miceDRF)

set.seed(17)
```

We will begin by loading the example dataset `windspeed` from the `mice` package:

```{r}
library(mice)

head(windspeed)

```

Next, we will introduce some missing values using the `ampute()` function:

```{r}
windspeed_miss <- ampute(windspeed, 0.2, mech  = "MAR")$amp
```

Finally, let’s check the proportion of missing values in each column:

```{r}
colMeans(is.na(windspeed_miss))
```


## Imputation Function

Before we can compute the *energy-I-Score*, we need to define the **imputation method** that will be applied to the dataset with missing values. The `miceDRF` package is flexible — it allows you to use **any imputation strategy**, as long as your function follows a few simple rules.

Your imputation function:

1. **Must accept** a dataset with missing values as its only required argument.  
2. **Must return** a dataset of the same dimensions, with missing values filled in.  
3. **Should not require any additional non-default arguments** — the function will be called internally by the scoring procedure.
4. Can be:
   - a simple custom function (e.g. mean imputation),
   - a wrapper around a more complex method (e.g. random forests, Bayesian models, `mice`, `missForest`),
   - or any other approach that returns a fully imputed dataset.

For example, let's define *zero imputation* below


```{r}
impute_zero <- function(X) { X[is.na(X)] <- 0; X }
```

Additionally, the `miceDRF` package provides a convenient way to quickly implement imputation methods from the `mice` package that are fully compatible with the energy-I-Score. You can simply use `create_mice_imputation()` with the name of the method (for the list of methods see `?mice::mice`) as below *random forest imputation*:

```{r}
impute_rf <- create_mice_imputation("rf")
```

## Energy-I-Score

To calculate Energy-I-Score we need to provide imputation method along with incomplete and imputed datasets. Let's see an example for random forest imputation:

```{r, warning=FALSE, message=FALSE}
sc <- Iscore(windspeed_miss, impute_rf(windspeed_miss), imputation_func = impute_rf)

sc
```

The result is a single score summarizing the performance across all columns that required imputation.
In addition, a table with scores calculated for each column separately is also returned as an attribute of the result. The table says what were score, weight and number of columns used for the training part for each column. To access this table, simply use the `attr()` function.

```{r}
attr(sc, "dat")
```

The `Iscore()` function exposes several parameters that control how the Energy-I-Score is computed.  
Below we illustrate the most important ones with concise examples.

The parameter `N` controls how many times the missing part is re-imputed to estimate the score (only relevant if your method is multiple). Let us note that for deterministic methods, `N` is effectively ignored (when `multiple = FALSE`).


```{r, warning=FALSE, message=FALSE}
Iscore(windspeed_miss, impute_rf(windspeed_miss), imputation_func = impute_rf, N = 5)
```
When working with datasets that contain many columns with missing values, calculating the energy-I-Score for all of them can be time-consuming.  To make the computation faster, you can **limit the number of columns used for scoring** by setting the `max_length` parameter. This parameter defines how many variables (columns) will be used to compute the score. The function automatically selects the variables **with the largest number of missing values first**. By default, `max_length = NULL`, which means that **all columns** with missing data are included in the calculation.


```{r, warning=FALSE, message=FALSE}
Iscore(windspeed_miss, impute_rf(windspeed_miss), imputation_func = impute_rf, max_length = 2)
```

Some variables may lack fully observed predictors among rows needed for training the conditional imputations. Thus, we can specify

- `skip_if_needed = TRUE` (default): try to skip minimal rows to obtain a workable design; proceed with scoring.

- `skip_if_needed = FALSE`: if no complete predictors can be formed, the score for that variable is returned as NA.

Additionally, we can adopt some scaling by `scale = TRUE` to ensure each variable is standardized internally before computing distances. This prevents variables with large numeric ranges from dominating the score.


### Summary of best practices

- Use `multiple = TRUE` with a genuinely multiple imputers and determine `N` for stable estimates.

- Consider `scale = TRUE` when mixing variables on different scales.

- Use `max_length` for quick experiments; remove it for final runs.

- Keep `skip_if_needed = TRUE` unless you explicitly want to flag unscorable columns with NA.


### Comparisons of energy-I-Scores

We also provide functionality for **quick benchmarking** of different imputation methods. To run Energy-I-Scores for more than one imputation function at once, use the `Iscores_compare()` function. You simply need to provide an incomplete dataset and a **list of imputation functions**, as shown below:

```{r, warning=FALSE, message=FALSE}
imputation_list <- list(rf = impute_rf, zero = impute_zero)

Iscores_compare(windspeed_miss, imputation_list, N = 10)
```

In the example above, we compare random forest imputation with a simple zero imputation method. According to the results, the random forest method yields a lower Energy-I-Score, indicating better imputation quality compared to the zero imputation baseline.

## Energy-I-Score for mixed datasets

When the data is mixed, i.e. it contains both categorical and numerical variables, you can use `IScore_cat()` function which is able to calculate scores on categorical columns. Before using it make sure that all the variables that contain categorical values are stored as **factors**.

## Energy score

If you have access to the original dataset before imputation, you can also use the energy distance as an additional evaluation metric. Our package provides an easy-to-use wrapper `energy_dist()` around the `energy::eqdist.e` function from the energy package. You can use it providing complete and imputed datasets:

```{r}
energy_dist(windspeed, impute_rf(windspeed_miss))

energy_dist(windspeed, impute_zero(windspeed_miss))
```

# References

This approach follows the methodology proposed by Näf, Grzesiak, and Scornet (2025) in "How to rank imputation methods?" (arXiv:2507.11297).









